---
title: "DATA 202 - Week 11"
subtitle: "On regression"
author: "Nathan Alexander, PhD"
institute: "Center for Applied Data Science and Analytics"
format: 
  html: default
  revealjs:
    output-file: week11-slides.html
    height: 900
    width: 1600
    smaller: false
    scrollable: true
    slide-number: c/t #< collapsted/total
    logo: "img/howard-logo.jpg"
    footer: "[Course Data GitHub](https://github.com/data-202)"
    toc: false
    echo: true
    incremental: false
    self-contained: true
---

## Part I: Context

This week we focus on more examples with added attention to our assumptions. We also explore the case of bivariate regression when attempting to measure and model conceptions of in/justice. Ahead of modeling multiple variables, we'll examine the importance of underlying theoretical analyses ahead of solely doing statistical investigations.

------------------------------------------------------------------------

Let us begin with a set of three variables: $x$, $y$, and $z$.

We will assume that there is a hypothesized association between all three variables.

However, let us also assume that we have *not* yet taken time to properly structure the relationships between the variables. Namely, we have yet to consider if $z$, for example, is a mediator, moderator, or spurious variable.

```{r, echo=T, message=F, warning=F}
## Sample outline of a bivariate regression analysis

# Generate sample data
set.seed(400)
x <- rnorm(100)
y <- 2 + -3*x^2 + rnorm(100)
z <- -4 + 3*x - rnorm(100)
df <- data.frame(x, y, z)
```

Let us take a look at the scatterplot between each pair of variables.

```{r, echo=T, message=F, warning=F}
plot(x, y)
cor(x, y)
```

Take note of the value of the correlation coefficient.

------------------------------------------------------------------------

Let us take a look at the scatterplot between the next pair of variables.

```{r, echo=T, message=F, warning=F}
plot(x, z)
cor(x, z)
```

Take note of the value of the correlation coefficient.

------------------------------------------------------------------------

Let us take a look at the scatterplots between each pair of variables.

```{r, echo=T, message=F, warning=F}
plot(y, z)
cor(y, z)
```

Take note of the value of the correlation coefficient.

------------------------------------------------------------------------

### Regression assumptions

In our previous lectures, we have discussed the function of assumptions and principles in regression analysis. In progressing towards different types of tests, it is important to consider the specific assumptions for any given test.

Analyzing relationships among social science variables has an assumptions of linearity. However, this assumptions is not always correct. The adoption of this assumption is based on a host of factors. Most notably, that many relationships have been found to be linear when considered in the empirical sense.

Some additional assumptions are as follows:

-   The sample is representative of the population

-   The variables of interest are normally distributed

-   There are no outliers in the data

-   Independence assumption

-   There is a linear relationship between the independent variable(s) and dependent variable(s)

------------------------------------------------------------------------

::: columns
::: {.column wideth="30%"}
- **Outcome, *Y***: The variable describing the outcome of interest.
- **Predictor, *X***: The variable used to help understand variability in the outcome.
:::

---

## Part II: Content

### A tradition: Fitting a straight line

In statistics, we often inquire about the relationship between two variables, $X$ and $Y$. These variables are sets that contain values (observations) as noted before. To do so, we use a regression model.

A **regression model** is a function that helps us describe how variables are related to one another. When we consider the simple case of a regression model, we take into account two variables, or a bivariate regression.

```{r, echo=T, message=F, warning=F}
library(tidyverse)
library(dplyr)
ggplot(data=df,
       mapping=aes(x = x, y = y)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  labs(
    x = "Independent variable",
    y = "Dependent variable"
  )
```
What do you notice? What do you wonder?

---

Let's now consider the relationship between x and z.


```{r, echo=T, message=F, warning=F}
library(tidyverse)
library(dplyr)
ggplot(data=df,
       mapping=aes(x = x, y = z)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  labs(
    x = "Independent variable",
    y = "Dependent variable"
  )
```

What do you notice? What do you wonder?

---

Now, let's look at the relationship between y and z.


```{r, echo=T, message=F, warning=F}
library(tidyverse)
library(dplyr)
ggplot(data=df,
       mapping=aes(x = y, y = z)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  labs(
    x = "Independent variable",
    y = "Dependent variable"
  )
```

------------------------------------------------------------------------

#### Matheamtics and exact relationships

In mathematics, two variables, $x$ and $y$ may be related to each other in various ways. These relationships form the basis for many inquiries in statistics. However, mathematical modeling most often involves identifying the parameters in which we can use to assess the validity of the stated relationship. In this way, we examine the difference between exact and inexact relationships.

------------------------------------------------------------------------

### A straight line

The equation of a line: $$y = mx + b$$

------------------------------------------------------------------------

#### Statistical modeling and inexact relationships

Regression analysis is a standard analysis in many statistical studies.

There are many different types of regression. We'll continue with our exploration of bivariate regression analysis and focus on some of the base assumptions as it relates to study development.

We will begin by looking at the underlying assumptions of regression analysis.

These assumptions are the technical (or structural) components of our analyses, and should be checked at the initiation of a research study, starting with data collection or understanding how data was collected if it is a secondary analysis.

------------------------------------------------------------------------

## Regression model {#regression-model-1}

A **regression model** is a function that describes the relationship between the outcome, $Y$, and the predictor, $X$.

$$\begin{aligned} Y &= \color{black}{\textbf{Model}} + \text{Error} \\[8pt]
&= \color{black}{\mathbf{f(X)}} + \epsilon \\[8pt]
&= \color{black}{\boldsymbol{\mu_{Y|X}}} + \epsilon \end{aligned}$$

## Regression model

::: columns
::: {.column width="30%"}
$$
\begin{aligned} Y &= \color{purple}{\textbf{Model}} + \text{Error} \\[8pt]
&= \color{purple}{\mathbf{f(X)}} + \epsilon \\[8pt]
&= \color{purple}{\boldsymbol{\mu_{Y|X}}} + \epsilon 
\end{aligned}
$$
:::

## Regression model + residuals

::: columns
::: {.column width="30%"}
$$\begin{aligned} Y &= \color{purple}{\textbf{Model}} + \color{blue}{\textbf{Error}} \\[8pt]
&= \color{purple}{\mathbf{f(X)}} + \color{blue}{\boldsymbol{\epsilon}} \\[8pt]
&= \color{purple}{\boldsymbol{\mu_{Y|X}}} + \color{blue}{\boldsymbol{\epsilon}} \\[8pt]
 \end{aligned}$$
:::

### Simple linear regression

We use simple linear regression to model the relationship between some predictor and outcome.

- $\beta_1$: True slope of the relationship between $X$ and $Y$
- $\beta_0$: True intercept of the relationship between $X$ and $Y$
- $\epsilon$: Error (residual)

$$\text{residual} = \text{observed} - \text{predicted} = y - \hat{y}$$

## Least squares line

-   The residual for the $i^{th}$ observation is

$$e_i = \text{observed} - \text{predicted} = y_i - \hat{y}_i$$

-   The **sum of squared** residuals is

$$e^2_1 + e^2_2 + \dots + e^2_n$$

-   The **least squares line** is the one that minimizes the sum of squared residuals

#### Properties of least squares regression

-   The regression line goes through the center of mass point, the coordinates corresponding to average $X$ and average $Y$: $\hat{\beta}_0 = \bar{Y} - \hat{\beta}_1\bar{X}$

-   The slope has the same sign as the correlation coefficient: $\hat{\beta}_1 = r \frac{s_Y}{s_X}$

-   The sum of the residuals is zero: $\sum_{i = 1}^n \epsilon_i = 0$

-   The residuals and $X$ values are uncorrelated

#### Estimating the slope

$$\large{\hat{\beta}_1 = r \frac{s_Y}{s_X}}$$

#### Estimating the intercept

$$\large{\hat{\beta}_0 = \bar{Y} - \hat{\beta}_1\bar{X}}$$

##### Analyzing two numeric variables

```{r}
#| echo: false
#| output: false
#| warning: false
## Sample outline of a bivariate regression analysis

# Generate sample data
set.seed(123)
x <- rnorm(100)
y <- 2*x*rnorm(100) + 3*x + rnorm(100)
```

Let us take two variables, $x$ and $y$.

```{r}
cor(x, y)
```

------------------------------------------------------------------------

```{r}
# Fit simple linear regression model
model <- lm(y ~ x)
```

------------------------------------------------------------------------

```{r}
# Examine regression outputs
summary(model)
```

------------------------------------------------------------------------

```{r}
# Check regression coefficients
coeff <- coef(model)
coeff
```

------------------------------------------------------------------------

```{r}
# Check R-squared value
rsq <- summary(model)$r.squared
rsq
```

------------------------------------------------------------------------

```{r}
# Generate predictions
pred <- predict(model) # call this object to show the predicted values of the model
```

------------------------------------------------------------------------

```{r}
# Plot the regression line
plot(x, y)
abline(model, col="blue")
```

------------------------------------------------------------------------

```{r}
# Check residuals
resids <- residuals(model)
plot(x, resids)
```

------------------------------------------------------------------------

```{r}
# Diagnostic plots
par(mfrow=c(2,2))
plot(model)
```

------------------------------------------------------------------------

```{r}
# Check significance of predictor
anova(model)
```

------------------------------------------------------------------------

#### Analyzing a categorical and a numeric variable

```{r}
#| echo: false
#| output: false
#| warning: false

require(dplyr)
require(ggplot2)
require(summarytools)

# Create a dataframe with race and income variables
df <- data.frame(
  race = c("Black", "White", "White", "Other", "Black", "Other", "White", "Black", "Other", "White"),
  income = c(25000, 35000, 42000, 31000, 27000, 38000, 29000, 40000, 32000, 39000)
)
```

------------------------------------------------------------------------

```{r}
desc(df$income)
```

------------------------------------------------------------------------

```{r}
# Bar plot of the categorical variable (race)
ggplot(df, aes(x = race)) +
  geom_bar(fill = "steelblue") +
  labs(x = "Race", y = "Count") +
  ggtitle("Distribution of Races")
```

```{r}
# Box plot of income by race
ggplot(df, aes(x = race, y = income)) +
  geom_boxplot(fill = "lightgray", color = "steelblue") +
  labs(x = "Race", y = "Income") +
  ggtitle("Income by Race")
```

#### Analyzing two categorical variables

In previous lectures, we discussed some methods to hypothesize around the relationship between two variables that are categorical in nature.

## Part III: Code

We will take a look at the `bivaraite.Rmd` file on Canvas.
