---
title: "Statistical Modeling Assumptions"
author: "Nathan Alexander"
output: html_document
---

## Introduction

This document introduces key concepts in modeling mediating, moderating, and spurious relationships among variables using the **gssr** package with GSS data. It also demonstrates how to test a simple path model.

## Mathematical Models and Definitions

### Mediation

Mediation occurs when the effect of an independent variable $X$ on a dependent variable $Y$ is transmitted through a mediator variable $M$. The total effect $c$ is decomposed into:

$$
c = c' + ab
$$

where\
- $a$ = effect of $X$ on $M$\
- $b$ = effect of $M$ on $Y$ controlling for $X$\
- $c'$ = direct effect of $X$ on $Y$ controlling for $M$

### Moderation

Moderation occurs when the effect of $X$ on $Y$ depends on the level of a moderator variable $W$. The model includes an interaction term:

$$
Y = \beta_0 + \beta_1 X + \beta_2 W + \beta_3 (X \times W) + \epsilon
$$

Significant $\beta_3$ indicates moderation.

### Spuriousness

A spurious relationship between $X$ and $Y$ occurs when a third variable $Z$ influences both, creating an apparent association. Controlling for $Z$ removes the effect of $X$ on $Y$.

## Setup and data loading

We begin by installing and loading the GSSR package.

```{r, message=F, warning=F}
install.packages("gssr")
library(gssr)

library(dplyr)
library(ggplot2)
library(haven)
library(tibble)
library(survey)
library(srvyr)
```

```{r, message=F, warning=F}

```

```{r, message=F, warning=F}

```

```{r, message=F, warning=F}

```

```{r, message=F, warning=F}

```

```{r, message=F, warning=F}

```

```{r, message=F, warning=F}

```

```{r, message=F, warning=F}

```

```{r, message=F, warning=F}

```
